Ab Initio Effective One-Electron Potential Operators. II. Applications for Exchange-Repulsion Energy in Effective Fragment Potentials

### 2019-12-12 papers/3_oep_technique.Pauli/paper-r0.tex
Data figures and tables added. Discussion completed

@@

On the other hand, it immediately follows that

@@

\begin{equation} \label{e:rep.murrell-etal.S1.oep-v}

@@

     = \sum_{x\in X}^{\rm At} W_{\zeta i}^{(x)} 

@@

Three complexes: 
(i) (H$_2$O)$_2$, 
(ii) H$_2$O--HOCH$_3$, 
(iii) (DMSO)$_2$,
were chosen as model systems to analyze the asymptotic dependence 
of EXR energy, which is shown in Figure~XXX. 
The reference (zero\hyp{}displacement) geometries
were obtained as described in Ref.\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019} 
and the structures  
along with the translation direction are depicted
in the insets in Figure~XXX.
Energy\hyp{}optimizations were performed at the HF/6-31+G(d,p) level,
as implemented in 
the {\sc Gaussian16} quantum chemistry program package.\cite{Gaussian16}
Small auxiliary basis sets for H$_2$O, CH$_3$OH and DMSO were developed
to minimize the computational cost of evaluating the summation terms
from Eq.~\eqref{e:rep.murrell-etal.S1.oep}. For this purpose, the basis set optimization 
as described in the Appendix~A in Ref.~\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019} was implemented
in conjunction with the basin hopping global optimization algorithm,  %cite
available in the SciPy Python module. %cite

To investigate the accuracy of the OEP\hyp{}based repulsion
term across
%perform statistical error analysis for 
a variety of
interacting systems, databases
of bi\hyp{}molecular complexes in the non\hyp{}covalent
interactions database NCB31 developed by the Truhlar's 
group,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
as well as the BioFragment Database subset BBI for backbone\hyp{}backbone
interactions in proteins,\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
as implemented in the {\sc Psi4}
program,\cite{Psi4.JCTC.2017}
were utilized as sources of equilibrium structures. 
In Figure~YYY we show statistical error analysis...

@@

Therefore, the following model is proposed for the
total exchange\hyp{}repulsion energy in the EFP2 model: 
(i) $E^{\rm Ex}$ and $E^{\rm Rep}(S^{-2})$ are evaluated
from the original EFP2 formulae of Jensen and Gordon, i.e.,
from Eqs.~\eqref{e:exc-efp2} and \eqref{e:rep-efp2.s2}, respectively;
(ii) $E^{\rm Rep}(S^{-1})$ term is evaluated from Eq.~\eqref{e:rep.murrell-etal.S1.oep}
and assuming a small auxiliary basis sets, optimized for each
EFP2 fragment separately.

>>>>

%
\begin{figure}[t]
\includegraphics[width=0.44\textwidth]{fig-1.pdf}
\caption{\label{f:fig-1} {\bf Asymptotic dependence of the exchange\hyp{}repulsion interaction energy
for selected bi\hyp{}molecular complexes.} 
%The charge transfer energy estimates are plotted for 
(a) water dimer, 
(b) water\hyp{}methanol complex, and 
(c) DMSO dimer in $S_i$ symmetry,
were one molecule has been translated
along the vector specified in the insets relative to initial geometry,
optimized at HF/6-31+G(d,p) level of theory.
The 
total interaction energy
is also shown for comparison in light grey color in this figure.
Interaction energies were obtained by using the 6-311++G(d,p)
primary basis set for (a) and (b), and 6-31++G(d,p) for (c).
In OEP calculations, EDF-1 scheme combined with small optimized auxiliary basis sets
(Supplementary Information) were used.
} 
\end{figure}
%

@@

{
\renewcommand{\arraystretch}{1.4}
\begin{table*}[t]
\caption[]
{{\bf CPU timings in miliseconds of exchange\hyp{}repulsion single point 
energy calculations via various methods\footnotemark[1]}
}
\label{t:tab-1}
\begin{ruledtabular}
\begin{tabular}{llccccccccc}
                      && \multicolumn{2}{c}{(H$_2$O)$_2$\footnotemark[2]} 
                      && \multicolumn{2}{c}{H$_2$O--HOCH$_3$\footnotemark[2]} 
                      && \multicolumn{2}{c}{(DMSO)$_2$\footnotemark[3]} \\
\cline{3-4}
\cline{6-7}
\cline{9-10}
Hayes \& Stone\footnotemark[4] && 1.16$\times 10^3$  & (4.81)  
                               && 1.12$\times 10^4$  & (4.72)
                               && 1.31$\times 10^5$  & (7.34) \\
Murrell et al.\footnotemark[5] && 2.33$\times 10^2$  & (4.53)  
                               && 3.87$\times 10^3$  & (4.44)
                               && 2.79$\times 10^4$  & (6.84) \\
EFP2                           && 0.496              & (4.45)  
                               && 0.84               & (4.44)
                               && 4.65   & (6.56) \\
OEP/large\footnotemark[6]      && 0.884  & (4.83)  
                               && 1.836  & (4.83)
                               && 8.95   & (6.77) \\
OEP/mini\footnotemark[6]       && 0.324  & (5.13)  
                               && 0.566  & (4.81)
                               && 3.00   & (7.41) \\
\end{tabular}
\end{ruledtabular}
\footnotetext[1]{1.2 GHz AMD EPYC\texttrademark{} 7301 16-Core Processor, calculations performed on 1 core. 
Exchange\hyp{}repulsion energies are given in parentheses for reference (kcal/mol).
Time profiling of the code performance was carried out as in Ref.~\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019}.
See also the implementation details in the main text.}
\footnotetext[2]{Primary basis set: 6-311++G(d,p).}
\footnotetext[3]{Primary basis set: 6-31++G(d,p).}
\footnotetext[4]{Reference~\cite{Hayes.Stone.MolPhys.1984}.}
\footnotetext[5]{Reference~\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965}.}
\footnotetext[6]{Eq.~\eqref{}. Auxiliary basis sets for the EDF in $E^{\rm Rep}(S^{-1})$ 
from Eq.~\eqref{e:rep.murrell-etal.S1.oep}
are: `large' - aug-cc-pVDZ-jkfit for (H$_2$O)$_2$ and H$_2$O--HOCH$_3$ systems as well as
aug-cc-pVDZ-ri for (DMSO)$_2$ system;
`mini' - optimized small basis sets (see Supporting Information).}
\end{table*}
}
\\ On the other hand, it immediately follows that

@@

\begin{multline} \label{e:rep.murrell-etal.S1.oep-v}

@@

     = \sum_{x\in X}^{\rm At} W_{\zeta i}^{(x)} \\

@@

We implemented EXR energy models from Eq.~\eqref{e:exr-efp2}
(EFP2), Eq.~\eqref{e:rep.murrell-etal.S1.oep} (OEP), as well as 
the approximate intermolecular perturbation theory with exchange of 
Murrell et~al.\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965},
%from Eq.~\eqref{e:rep.murrell-etal.S1} 
the exact EXR energy of Hayes \& Stone\cite{Hayes.Stone.MolPhys.1984}
and the density decomposition scheme 
at the Hartree\hyp{}Fock level of theory (here referred to as the DDS/HF) 
developed by Mandado and Hermida\hyp{}Ram{\'o}n,\cite{Mandado.Hermida-Ramon.JCTC.2011} 
for validation of the EDF technique.
When implemented in the dimer\hyp{}centered basis set\cite{Chalasinski.Gutowski.MolPhys.1985} (DCBS),
the DDS/HF model is
equivalent to the exact EXR energy by Hayes \& Stone and gives identical results,
which is the case in this work. Using DCBS also allows one to eliminate the basis set
superimposition error (DCBS).
To localize molecular orbitals, procedure that is necessary for the EFP2 calculations,
the Boys method\cite{Boys.RevModPhys.1960} was used. Note here that, since
Murrell et~al.'s theory is invariant with respect to unitary transformation
of molecular orbitals, the OEP\hyp{}based expression 
in Eq.~\eqref{e:rep.murrell-etal.S1.oep} is also invariant and does not require
orbital localization. 
%in which
%
%\begin{equation}\label{e:dds-exr}
% \Delta E^{\rm Ex-Rep}_{\rm DDS/HF} = 
%\end{equation}
%
%
\begin{figure*}[t]
\includegraphics[width=0.8\textwidth]{fig-2.pdf}
\caption{\label{f:fig-2} {\bf Accuracy of the OEP and EFP2 models of exchange\hyp{}repulsion energy
across various bi\hyp{}molecular systems.} 
(a) NCB31 
database,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005} 
of non\hyp{}covalent interactions
and
(b) BBI subset\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
of backbone\hyp{}backbone interactions in proteins from the BioFragment Database.
For the OEP calculations, the EDF-1 scheme with the aug-cc-pVDZ-jkfit auxiliary basis set
was used.
} 
\end{figure*}
%

Three complexes: 
(H$_2$O)$_2$, 
H$_2$O--HOCH$_3$, and
(DMSO)$_2$,
were chosen as model systems to analyze the asymptotic dependence 
of EXR energy, which is shown in Figure~\ref{f:fig-1}. 
The reference (zero\hyp{}displacement) geometries
were obtained as described in Ref.\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019} 
and the structures  
along with the translation direction are depicted
in the insets in Figure~\ref{f:fig-1}.
Energy\hyp{}optimizations were performed at the HF/6-31+G(d,p) level,
as implemented in 
the {\sc Gaussian16} quantum chemistry program package.\cite{Gaussian16}
OEP and EFP2 models correctly describe the EXR energy at all separations
for the studied model systems with very similar accuracy. 
However,
contrary to our previous application of the OEP technique\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019},
where approximately 20\hyp{}fold speedups as compared to the EFP2 model were achieved
for the evaluation of the charge\hyp{}transfer energy,
the computational cost of the OEP EXR method is roughly 2 times higher than the EFP2 model
when the usual density fitting auxiliary basis sets such as aug-cc-pVDZ-jkfit or aug-cc-pVDZ-ri are used
(Table~\ref{t:tab-1}).
However, application of our small auxiliary basis sets for H$_2$O, CH$_3$OH and DMSO, 
%developed
developed via the basis set optimization~\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019}
in conjunction with the basin hopping global optimization algorithm  %cite
as implemented in the SciPy Python library
(see Supplementary Information),
%to minimize the computational cost of evaluating the summation terms
%from Eq.~\eqref{e:rep.murrell-etal.S1.oep},
reduces costs appreciably and makes the OEP method on average 1.5 times faster than EFP2.
For instance, calculation of the EXR energy in the DMSO dimer by using 6-31++G(d,p) primary
basis set and 
1.2 GHz AMD EPYC\texttrademark{} 7301 16-Core Processor
requires about 4.6~ms and 3.0~ms when EFP2 and OEP model is used, respectively.
%Note that the benchmark Hayes \& Stone's calculations predict slightly larger repulsion energies

To investigate the accuracy of the OEP\hyp{}based repulsion
term across
%perform statistical error analysis for 
a variety of
interacting systems, range of
bi\hyp{}molecular complexes from the non\hyp{}covalent
interactions database NCB31 developed by the Truhlar's 
group,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
as well as the BioFragment Database subset BBI for backbone\hyp{}backbone
interactions in proteins of Sherrill group,\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
as implemented in the {\sc Psi4}
program,\cite{Psi4.JCTC.2017}
were utilized, and the
EXR energies were compared with the reference DDS/HF results (Figure~\ref{f:fig-2}).
On average, OEP and EXR tend to underestimate the EXR energy by 0.7\% and 7.6\%
in the NCB31/6-311++G(d,p) data set, and 6.0\% and 15\% in the BBI/6-31+G(d,p) data set,
respectively.
This is mostly due to the lack of BSSE, that is accounted for
only in the reference calculations with the DDS/HF model, and the effect is usually larger
when smaller primary basis sets are used.
The OEP model is of slightly better accuracy as the EFP2 model
in systems sudied in this work, with 
root mean square errors of OEP and EFP2 model being
2.0 kcal/mol and 2.6 kcal/mol for NCB31/6-311++G(d,p) data set, 
as well as
0.5 kcal/mol and 1.2 kcal/mol for BBI/6-31+G(d,p) data set, respectively.
It is an accidental fruitful cancellation of errors because
the first\hyp{}order repulsive terms have much less differences with respect to each other
(RMSE is 0.7\% only in all data sets).
Nevertheless, together with the performance data of both models
from Table~\ref{t:tab-1}, OEP model performs better than EFP2 as long as the minimal
auxiliary basis sets are used.

@@

In this work, the OEP technique of effective ERI elimination
proposed by us previously was utilized to reduce the computational cost
of evaluation of the EXR energy in the EFP2 model.
Starting from the Murrell et~al.'s model, which is a foundation of the EFP2 EXR term,
OEP\hyp{}based expression was derived which requires evaluation of only overlap
OEI's, apart from additional kinetic energy OEI's required for EFP2. 
Therefore, the following model is proposed for the
total exchange\hyp{}repulsion energy in the EFP2 model: 
(i) $E^{\rm Ex}$ and $E^{\rm Rep}(S^{-2})$ are evaluated
from the original EFP2 formulae of Jensen and Gordon, i.e.,
from Eqs.~\eqref{e:exc-efp2} and \eqref{e:rep-efp2.s2}, respectively;
(ii) $E^{\rm Rep}(S^{-1})$ term is evaluated from Eq.~\eqref{e:rep.murrell-etal.S1.oep}
and assuming a small auxiliary basis sets, optimized for each
EFP2 fragment separately.
